---
title: "4th Floor Fantast Draft 2016"
author: "Evan Oman"
date: "September 20, 2016"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation

First we load in the spreadsheet Jacob used to record the draft:

```{r}
library("reshape2")
library("plyr")
    
draftData <- read.csv("./Draft.csv")
print.data.frame(head(draftData), row.names=FALSE)
```

As you can see, the data is organized by round with a column for each drafter. This format is often refered to as a "wide dataset." For this analysis we want a "tall dataset" so we will use the "melt" fundtion to reorganize the data into one row per datapoint (ie one row per pick).

```{r warning=FALSE}
# Melt the data
draftData <- melt(draftData, id.vars = "ROUND")

# Drop the row names
rownames(draftData) <- NULL

# Order by round (this is a stable sort)
draftData <- draftData[order(draftData$ROUND), ]

# Rename some columns
draftData <- rename(draftData, c("variable" = "Drafter", "value" = "Player", "ROUND" = "Round"))

print.data.frame(head(draftData, n=10), row.names=FALSE)
```
Now that we have the data in a tall form, we want to know what the actual number for each pick. This is complicated by the winding nature of the draft order (Chris had the 8th and 9th picks). To capture this behavior I have written the following "snake function" which creates a sequence which winds up to 8 and then counts down from 16, up to 24, etc.
```{r}
# Creates the sequence: 1,2,3,4,5,6,7,8,16,15,14,13,12,11,10,9,17,18,19,20....
snakeFunction <- function(n, rowLen=8)
{
  lapply(1:n, function(i){
    groupIndicator <- ceiling(i / rowLen)
    # Increasing group
    if (groupIndicator %% 2 == 1)
    {
      i
    }
    # Decreasing group
    else
    {
      rowLen*groupIndicator - ((i - 1) %% rowLen)
    }
  })
}

# Create the list for out draft (8 drafters, 16 rounds)
pickOrdering <- snakeFunction(8*16)

unlist(pickOrdering)
```

We can then use this pick ordering list to index the draft picks in our draft data frame.

```{r}
# Add the pick numbers to the draft data
draftData$PickNumber <- pickOrdering
draftData <- as.data.frame(lapply(draftData, unlist))

# Order the draft data by pick number
draftData <- draftData[order(draftData$PickNumber),]

print.data.frame(head(draftData, n=10), row.names=FALSE)
```

Now that the draft data has been prepped, we can load in ESPN rankings (from espn.com)

```{r}
espn <- read.csv("./ESPN_Rankings.csv")

print.data.frame(head(espn, n=10), row.names=FALSE)
```

This data set is already in the format we want so we can go ahead and join it with our draft data with a match by Player. Note that this is a left join by default so the resulting dataframe will only contain rows from our draft dataframe. Some players will have will not have an ESPN ranking. After the join these players will have a `NA` rank, we will replace this with the worst possible ranking (`nrow(espn)`) plus `1`.

```{r}
# Join espn rankings data with draft data 
joinedData <- join(draftData, espn, by="Player")

# Some players will not have an ESPN ranking, give these players a worst ranking + 1
joinedData$Rank[is.na(joinedData$Rank)] <- nrow(espn) + 1

print.data.frame(head(joinedData, n=10), row.names=FALSE)
```

We now have a dataframe with each draft pick and the corresponding pick number and player ranking. We can use the pick number and player ranking to measure the quality of each draft pick.

```{r}
joinedData$RankDiff <- joinedData$PickNumber - joinedData$Rank

print.data.frame(head(joinedData, n=10), row.names=FALSE)
```

At this point all of the data has been prepped so we are ready to generate some plots and do a bit of analysis.

## Figures and Results

You can also embed plots, for example:

```{r, echo=FALSE}
library(ggplot2)
avgDiffs <- aggregate(RankDiff ~ Drafter, joinedData, FUN=mean)

avgDiffs <- within(avgDiffs, Drafter <- factor(Drafter, levels=avgDiffs[order(avgDiffs$RankDiff), "Drafter"]))

ggplot(avgDiffs, aes(x=Drafter, y=RankDiff)) + geom_bar(stat = "identity") + ylab("Difference Between Pick # and Ranking")
```

```{r, echo=FALSE}
posDiffs <- aggregate(RankDiff ~ Pos, joinedData, FUN=mean)

posDiffs <- within(posDiffs, Pos <- factor(Pos, levels=posDiffs[order(posDiffs$RankDiff), "Pos"]))

# Bar Plot
ggplot(posDiffs, aes(x=Pos, y=RankDiff)) + geom_bar(stat = "identity") + ylab("Difference Between Pick # and Ranking")
```

```{r, echo=FALSE}
# Scatter colored by drafter 
ggplot(joinedData, aes(x=PickNumber,y=Rank)) + geom_point(aes(color=Drafter)) + geom_abline() + xlab("Pick #")
```

```{r, echo=FALSE}
# Scatter colored by position 
ggplot(joinedData, aes(x=PickNumber,y=Rank)) + geom_point(aes(color=Pos)) + geom_abline() + xlab("Pick #")
```

```{r, echo=FALSE}
# Scatter colored by position 
ggplot(joinedData, aes(x=Round, y=RankDiff)) + geom_line() + facet_wrap(~Drafter, ncol=2) + geom_hline(yintercept = 0, linetype=2) + geom_point(aes(color=Pos)) + ylab("Difference Between Pick # and Ranking")
```
