---
title: "4th Floor Fantast Draft 2016"
author: "Evan Oman"
date: "September 20, 2016"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
options(width=100)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Preparation

First we load in the spreadsheet Jacob used to record the draft:

```{r}
library("reshape2")
library("plyr")
    
draftData <- read.csv("./Draft.csv")
print.data.frame(head(draftData), row.names=FALSE)
```

As you can see, the data is organized by round with a column for each drafter. This format is often referred to as a "wide data-set." For this analysis we want a "tall data set" so we will use the "melt" function to reorganize the data into one row per data point (ie one row per pick).

```{r warning=FALSE}
# Melt the data
draftData <- melt(draftData, id.vars = "ROUND")

# Drop the row names
rownames(draftData) <- NULL

# Order by round (this is a stable sort)
draftData <- draftData[order(draftData$ROUND), ]

# Rename some columns
draftData <- rename(draftData, c("variable" = "Drafter", "value" = "Player", "ROUND" = "Round"))

print.data.frame(head(draftData, n=10), row.names=FALSE)
```
Now that we have the data in a tall form, we want to know what the actual number for each pick. This is complicated by the winding nature of the draft order (Chris had the 8th and 9th picks). To capture this behavior I have written the following "snake function" which creates a sequence which winds up to 8 and then counts down from 16, up to 24, etc.
```{r}
# Creates the sequence: 1,2,3,4,5,6,7,8,16,15,14,13,12,11,10,9,17,18,19,20....
snakeFunction <- function(n, rowLen=8)
{
  lapply(1:n, function(i){
    groupIndicator <- ceiling(i / rowLen)
    # Increasing group
    if (groupIndicator %% 2 == 1)
    {
      i
    }
    # Decreasing group
    else
    {
      rowLen*groupIndicator - ((i - 1) %% rowLen)
    }
  })
}

# Create the list for out draft (8 drafters, 16 rounds)
pickOrdering <- snakeFunction(8*16)

unlist(pickOrdering)
```

We can then use this pick ordering list to index the draft picks in our draft data frame.

```{r}
# Add the pick numbers to the draft data
draftData$PickNumber <- pickOrdering
draftData <- as.data.frame(lapply(draftData, unlist))

# Order the draft data by pick number
draftData <- draftData[order(draftData$PickNumber),]

print.data.frame(head(draftData, n=10), row.names=FALSE)
```

Now that the draft data has been prepped, we can load in ESPN rankings (from espn.com)

```{r}
espn <- read.csv("./ESPN_Rankings.csv")

print.data.frame(head(espn, n=10), row.names=FALSE)
```

This data set is already in the format we want so we can go ahead and join it with our draft data with a match by Player. Note that this is a left join by default so the resulting dataframe will only contain rows from our draft dataframe. Note that some players will have will not have an ESPN ranking so after the join these players will have a `NA` rank. We will replace this with the worst possible ranking (`nrow(espn)`) plus `1`.

```{r}
# Join espn rankings data with draft data 
joinedData <- join(draftData, espn, by="Player")

# Some players will not have an ESPN ranking, give these players a worst ranking + 1
joinedData$Rank[is.na(joinedData$Rank)] <- nrow(espn) + 1

print.data.frame(head(joinedData, n=10), row.names=FALSE)
```

We now have a dataframe with each draft pick and the corresponding pick number and player ranking. We can use the pick number and player ranking to measure the quality of each draft pick.

```{r}
joinedData$RankDiff <- joinedData$Rank - joinedData$PickNumber

print.data.frame(head(joinedData, n=10), row.names=FALSE)
```

At this point all of the data has been prepped so we are ready to generate some plots and do a bit of analysis.

## Figures and Results

Probably the most natural measure of performance would be to look at the average difference between pick number and rank, across all rounds, organized by Drafter. We can calculate this using `aggregate`.

```{r}
avgDiffs <- aggregate(RankDiff ~ Drafter, joinedData, FUN=mean)

# Order factor by worst draft to best
avgDiffs <- within(avgDiffs, Drafter <- factor(Drafter, levels=avgDiffs[order(avgDiffs$RankDiff), "Drafter"]))

# Order dataframe by worst draft to best
avgDiffs <- avgDiffs[order(avgDiffs$RankDiff, decreasing = FALSE),]

print.data.frame(avgDiffs, row.names=FALSE)
```

Thus we can see that Chris had the best draft (on average, he got players 11.8 positions under rank value) and Kuuuby had the worst (on average, he got players 80.5 positions over rank value). Here is a quick visual of the above dataframe.

```{r}
library(ggplot2)
ggplot(avgDiffs, aes(x=Drafter, y=RankDiff)) + geom_bar(stat = "identity") + ylab("Difference Between Pick # and Ranking")
```

I was curious to see how undervalued/overvalued players were by position. We can use a similar methodology as above to figure this out.

```{r}
posDiffs <- aggregate(RankDiff ~ Pos, joinedData, FUN=mean)

posDiffs <- within(posDiffs, Pos <- factor(Pos, levels=posDiffs[order(posDiffs$RankDiff), "Pos"]))

# Order dataframe by worst draft to best
posDiffs <- posDiffs[order(posDiffs$RankDiff, decreasing = FALSE),]

print.data.frame(posDiffs, row.names=FALSE)

ggplot(posDiffs, aes(x=Pos, y=RankDiff)) + geom_bar(stat = "identity") + ylab("Difference Between Pick # and Ranking")
```

Here we can see that on average, RBs were undervalued by 6.6 rankings and D/STs were overvalued by 82 rankings.

I was also curious to see how draft pick quality changed by round. In the next two figures I am plotting `Pick #` vs `Rank` with a diagonal line indicating a pick at the ESPN rank value. Thus all picks below this line are "good" and all of the above are "bad". In the first figure I colored the picks by Drafter and in the second I am coloring by position.

```{r}
# Scatter colored by drafter 
ggplot(joinedData, aes(x=PickNumber,y=Rank)) + geom_point(aes(color=Drafter)) + geom_abline() + xlab("Pick #")
```

```{r}
# Scatter colored by position 
ggplot(joinedData, aes(x=PickNumber,y=Rank)) + geom_point(aes(color=Pos)) + geom_abline() + xlab("Pick #")
```

It is pretty clear that as time goes on, worse and worse picks made.

Finally I wanted to see how each drafter performed by round, the following figure accomplishes this.

```{r}
# Scatter colored by position, faceted by drafer
ggplot(joinedData, aes(x=Round, y=RankDiff)) + geom_line() + facet_wrap(~Drafter, ncol=2) + geom_hline(yintercept = 0, linetype=2) + geom_point(aes(color=Pos)) + ylab("Difference Between Pick # and Ranking")
```

## Conclusion
Here are some final take aways:

- **Best Draft:** Chris (avg pick 11.8 positions under rank value)
- **Worst Draft:** Kuuby (avg pick 80.5 positions over rank value)
- **Best Pick:** Drafter: `r p <- joinedData[which.min(joinedData$RankDiff),]; p$Drafter`, Round: `r p$Round`, Player: `r p$Player`,  Rank Difference: `r p$RankDiff`
- **Worst Pick:** Drafter: `r p <- joinedData[which.max(joinedData$RankDiff),]; p$Drafter`, Round: `r p$Round`, Player: `r p$Player`,  Rank Difference: `r p$RankDiff`